\documentclass{../office-sheet}

\begin{document}

%% TODO: correct date
\chapter*{Matins of the Dead}

 \textit{On All Souls' Day, all three Nocturns are said. On other days, only one Nocturn is said according to the following order: first Nocturn on Sundays, Mondays, and Thursdays; second Nocturn on Tuesdays and Fridays; third Nocturn on Wednesdays and Saturdays. On the day of burial, the first Nocturn is always said.}

%% TODO: correct the little chapter
 \section*{The Invitatory}
 
%\gresetinitiallines{1}
%\gregorioscore{matins_element_1}

\textit{Come, let us adore the King, * to whom all things live.}

\gresetinitiallines{1}
\gregorioscore{matins_element_1}

{\color{red}\textit{The Choir repeats:} Regem cui ómnia vivunt, * Veníte adorémus.}

 \section*{Psalm 94}

\gresetinitiallines{1}
\gregorioscore{matins_element_2}

\gresetinitiallines{1}
\gregorioscore{matins_element_3}

\gresetinitiallines{1}
\gregorioscore{matins_element_4}

\gresetinitiallines{1}
\gregorioscore{matins_element_5}

\gresetinitiallines{1}
\gregorioscore{matins_element_6}

\gresetinitiallines{1}
\gregorioscore{matins_element_7}

\textit{Psalm 94: (1) Come, let us praise the Lord with joy; let us joyfully sing to God our Savior. Let us come before His presence with thanksgiving, and make a joyful noise to Him with psalms. (a) Come, let us adore the King, to whom all things live. (2) For the Lord is a great God, and a great king above all gods. For the Lord will not reject His people. For in His hands are all the ends of the earth, and the heights of the montains are His. (a) Come, let us adore. (3) For the sea is His, and He made it, and His hands formed the dry land. Come, let us adore and fall down, and weep before the Lord that made us. For He is the Lord oour God: and we are His people, and the sheep of His pasture. (a) Come, let us adore the King, to whom all things live. (4) Today, if you shall hear His voice, harden not your hearts. As in the provocation, according to the day of temptation in the wilderness: where your fathers tempted Me, they proved Me, and saw My works. (a) Come, let us adore. (5) Forty years long I was offended with that generation, and I said: These always err in heart. And these men have not known My ways: so I swore in My wrath, that they shall not enter into My rest. (a) Come, let us adore the King, to whom all things live. (6) Eternal rest grant unto them, O Lord, * and let perpetual line shine upon them. (a) Come, let us adore. (a) Come, let us adore the King, to whom all things live.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

 \section*{First Nocturn}
 \begin{center}
 \textit{To be said on Sundays, Mondays, and Thursdays.}
 \end{center}
 
 \textit{Direct, O Lord, my God, my steps in Thy sight.}
 \gresetinitiallines{1}
 \gregorioscore{antiphon_dirige_domine}
 
 \subsection*{Psalm 5}
 
 \gresetinitiallines{1}
 \gregorioscore{intonation_psalm_5}
 
 \begin{latinenglishsection} 
 \latinenglish{
 	2. Inténde voci orati\textbf{ó}nis \textbf{me}æ:~* Rex meus et \textbf{De}us \textbf{me}us.

	3. Quóniam ad \textbf{te} o\textbf{rá}bo:~* Dómine, mane exáudies \textbf{vo}cem \textbf{me}am.
	
	4. Mane astábo tibi \textbf{et} vi\textbf{dé}bo:~* quóniam non Deus volens iniqui\textbf{tá}tem \textbf{tu} es.
	
	5. Neque habitábit juxta \textbf{te} ma\textbf{lí}gnus:~* neque permanébunt injústi ante \textbf{ó}culos \textbf{tu}os.
	
	6. Odísti omnes, qui operántur in\textbf{i}qui\textbf{tá}tem:~* perdes omnes, qui lo\textbf{quún}tur men\textbf{dá}cium.
	
	7. Virum sánguinum et dolósum abomi\textbf{ná}bitur \textbf{Dó}minus:~* ego autem in multitúdine miseri\textbf{cór}diæ \textbf{tu}æ.
	
	8. Introíbo in \textbf{do}mum \textbf{tu}am:~* adorábo ad templum sanctum tuum in ti\textbf{mó}re \textbf{tu}o.
	
	9. Dómine, deduc me in jus\textbf{tí}tia \textbf{tu}a:~* propter inimícos meos dírige in conspéctu tuo \textbf{vi}am \textbf{me}am.
	
	10. Quóniam non est in ore e\textbf{ó}rum \textbf{vé}ritas:~* cor e\textbf{ó}rum \textbf{va}num est.
	
	11. Sepúlcrum patens est guttur eórum,~{\color{red}\GreDagger} linguis suis do\textbf{ló}se a\textbf{gé}bant,~* júdica \textbf{il}los, \textbf{De}us.
	
	12. Décidant a cogitatiónibus suis,~{\color{red}\GreDagger} secúndum multitúdinem impietátum eórum ex\textbf{pél}le \textbf{e}os,~* quóniam irrita\textbf{vé}runt te, \textbf{Dó}mine.
	
	13. Et læténtur omnes, qui \textbf{spe}rant \textbf{in} te,~* in ætérnum exsultábunt: et habi\textbf{tá}bis in \textbf{e}is.
	
	14. Et gloriabúntur in te omnes, qui díligunt \textbf{no}men \textbf{tu}um:~* quóniam tu bene\textbf{dí}ces \textbf{jus}to.
	
	15. Dómine, ut scuto bonæ volun\textbf{tá}tis \textbf{tu}æ~* \textbf{co}ro\textbf{nás}ti nos.
	
	16. Requiem æternam dona \textbf{e}is, \textbf{Do}mine~* Et lux perpetua lu\textbf{ce}at \textbf{e}is.
 }{
 	1. Give ear, O Lord, to my words, understand my cry.
 	
 	2. Hearken to the voice of my prayer, O my King and my God.
 	
 	3. For to Thee will I pray: O Lord, in the morning Thou shalt hear my voice.
 	
 	4. In the morning I shall stand before Thee, and will see; because Thou art not a God that willeth iniquity.
 	
 	5. Neither shall the wicked dwell near Thee: nor shall the unjust abide before They eyes.
 	
 	6. Thou hatest all workers of iniquity: Thou wilt destroy all that speak a lie.
 	
 	7. The Lord abhorreth teh bloody and deceitful man. But as for me, in the multitude of Thy mercy,
 	
 	8. I will come into Thy house; I will adore, in Thy fear, in Thy holy temple.
 	
 	9. Conduct me, O Lord, in Thy justice: because of my enemies, direct my way in Thy sight.
 	
 	10. For there is no truth in their mouth: their heart is vain.
 	
 	11. Their throat is an open sepulchre: with their tongues they deal deceitfully: judge them, O God.
 	
 	12. Let them fall from their devices: according to the multitude of their iniquities cast them out: for they have provoked Thee, O Lord.
 	
 	13. But let all that hipe in Thee be glad: they shall rejoice forever, and Thou shalt dwell in them.
 	
 	14. And all they that love Thy name shall glory in Thee, for Thou wilt bless the just.
 	 	
 	15. O Lord, Thou hast crowned us with Thy good will as with a shield.
 	
 	16. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}
 
\gresetinitiallines{1}
\gregorioscore{antiphon_dirige_domine}
\textit{Direct, O Lord, my God, my steps in Thy sight.}
 
%%

\hspace{2cm}
 
\textit{Turn, O Lord, and deliver my soul, for there is none in death who will be mindful of Thee.}
\gresetinitiallines{1}
\gregorioscore{antiphon_convertere_domine}
 
\subsection*{Psalm 6}
 
\gresetinitiallines{1}
\gregorioscore{intonation_psalm_6}
 
\begin{latinenglishsection} 
\latinenglish{
 	2. Miserére mei, Dómine, quóniam in\textbf{fír}mus sum:~* sana me, Dómine, quóniam conturbáta sunt \textit{os}\textit{sa} textbf{me}a.

	3. Et ánima mea turbáta est \textbf{val}de:~* sed tu, Dó\textit{mi}\textit{ne}, \textbf{ús}quequo?

	4. Convértere, Dómine, et éripe ánimam \textbf{me}am:~* salvum me fac propter misericór\textit{di}\textit{am} \textbf{tu}am.
	
	5. Quóniam non est in morte qui memor sit \textbf{tu}i:~* in inférno autem quis confité\textit{bi}\textit{tur} \textbf{ti}bi?
	
	6. Laborávi in gémitu meo,~{\color{red}\GreDagger} lavábo per síngulas noctes lectum \textbf{me}um:~* lácrimis meis stratum me\textit{um} \textit{ri}\textbf{gá}bo.
	
	7. Turbátus est a furóre óculus \textbf{me}us:~* inveterávi inter omnes ini\textit{mí}\textit{cos} \textbf{me}os.
	
	8. Discédite a me, omnes, qui operámini iniqui\textbf{tá}tem:~* quóniam exaudívit Dóminus vocem \textit{fle}\textit{tus} \textbf{me}i.
	
	9. Exaudívit Dóminus deprecatiónem \textbf{me}am,~* Dóminus oratiónem me\textit{am} \textit{su}\textbf{scé}pit.
	
	10. Erubéscant, et conturbéntur veheménter omnes inimíci \textbf{me}i:~* convertántur et erubéscant val\textit{de} \textit{ve}\textbf{ló}citer.
	
	11. Requiem æternam dona eis, \textbf{Do}mine~* Et lux perpetua lu\textit{ceat} \textbf{e}is.
 }{
 	1. O Lord, rebuke me not in Thy anger, nor chastise me in Thy wrath.
 	
 	2. Have mercy on me, O Lord, for I am weak: heal me, O Lord, for my bones are sore.
 	
 	3. And my soul is troubled exceedingly: but Thou, O Lord, how long?
 	
 	4. Turn, O Lord, and deliver my soul: O save me for Thy mercy's sake.
 	
 	5. For in death there is no one that is mindful of Thee: and who shall praise Thee in hell?
 	
 	6. I have labored in my groanings, every night I wash my bed: I water my couch with my tears.
 	
 	7. My eye is troubled through indignation: I have grown old amidst all my enemies.
 	
 	8. Depart from me, all ye workers of iniquity, for the Lord hath heard the voice of my weeping.
 	
 	9. The Lord hath heard my supplication: the Lord hath received my prayer.
 	
 	10. Let all my enemies be put to shame, and be very much troubled: let them be put to shame very speedily.
 	
 	16. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}
 
\gresetinitiallines{1}
\gregorioscore{antiphon_convertere_domine}
\textit{Turn, O Lord, and deliver my soul, for there is none in death who will be mindful of Thee.}
 
 %%
 
\hspace{2cm}
 
\textit{Lest at any time, like a lion, he seize upon my soul, while there is no one to rescue me, nor to save.}
\gresetinitiallines{1}
\gregorioscore{antiphon_nequando_rapiat}
 
\subsection*{Psalm 7}
 
\gresetinitiallines{1}
\gregorioscore{intonation_psalm_7}

\begin{latinenglishsection} 
\latinenglish{
 	2. Nequándo rápiat ut leo ánimam \textbf{me}am,~* dum non est qui rédimat, neque qui \textit{sal}\textit{vum} \textbf{fá}ciat.

	3. Dómine, Deus meus, si feci \textbf{is}tud.~* si est iníquitas in má\textit{ni}\textit{bus} \textbf{me}is:
	
	4. Si réddidi retribuéntibus mihi \textbf{ma}la,~* décidam mérito ab inimícis me\textit{is} \textit{in}\textbf{á}nis.
	
	5. Persequátur inimícus ánimam meam,~{\color{red}\GreDagger} et\\ comprehéndat, et concúlcet in terra vitam \textbf{me}am,~* et glóriam meam in púlve\textit{rem} \textit{de}\textbf{dú}cat.
	
	6. Exsúrge, Dómine, in ira \textbf{tu}a:~* et exaltáre in fínibus inimicó\textit{rum} \textit{me}\textbf{ó}rum.
	
	7. Et exsúrge, Dómine Deus meus, in præcépto quod man\textbf{dás}ti:~* et synagóga populórum cir\textit{cúm}\textit{da}\textbf{bit} te.
	
	8. Et propter hanc in altum re\textbf{gré}dere:~* Dóminus jú\textit{di}\textit{cat} \textbf{pó}pulos.
	
	9. Júdica me, Dómine, secúndum justítiam \textbf{me}am,~* et secúndum innocéntiam \textit{me}\textit{am} \textbf{su}per me.
	
	10. Consumétur nequítia peccatórum, et díriges \textbf{jus}tum,~* scrutans corda et \textit{re}\textit{nes} \textbf{De}us.
	
	11. Justum adjutórium meum a \textbf{Dó}mino,~* qui salvos facit \textit{rec}\textit{tos} \textbf{cor}de.
	
	12. Deus judex justus, fortis, et \textbf{pá}tiens:~* numquid iráscitur per sín\textit{gu}\textit{los} \textbf{di}es?
	
	13. Nisi convérsi fuéritis, gládium suum vi\textbf{brá}bit:~* arcum suum teténdit, et pa\textit{rá}\textit{vit} \textbf{il}lum.
	
	14. Et in eo parávit vasa \textbf{mor}tis:~* sagíttas suas ardénti\textit{bus} \textit{ef}\textbf{fé}cit.
	
	15. Ecce partúriit injus\textbf{tí}tiam:~* concépit dolórem, et péperit in\textit{i}\textit{qui}\textbf{tá}tem.
	
	16. Lacum apéruit, et effódit \textbf{e}um:~* et íncidit in fóve\textit{am} \textit{quam} \textbf{fe}cit.
	
	17. Convertétur dolor ejus in caput \textbf{e}jus:~* et in vérticem ipsíus iníquitas e\textit{jus} \textit{de}\textbf{scén}det.
	
	18. Confitébor Dómino secúndum justítiam \textbf{e}jus:~* et psallam nómini Dómi\textit{ni} \textit{al}\textbf{tís}simi.
	
	19. Requiem æternam dona eis, \textbf{Do}mine~* Et lux perpetua lu\textit{ceat} \textbf{e}is.
 }{
 	1. O Lord my God, in Thee have I put my trust: save me and deliver me from all them that persecute me.
 	
 	2. Lest at any time, like a lion, he seize upon my soul, while there is no one to rescue me, or to save.
 	
 	3. O Lord, my God, if I have done this thing, if there be iniquity in my hands:
 	
 	4. If I have made a return to them that did evils to me, let me deservedly fall powerless before my enemies.
 	
 	5. Let the enemy pursue my soul, and take it, and tread down my life on the earth, and bring down my glory to the dust.
 	
 	6. Rise up, O Lord, in Thy anger: and be Thou exalted in the borders of my enemies.
 	
 	7. And arise, O Lord, my God, for the judgement which Thou hast commanded: and a congregation of people shall surround Thee.
 	
 	8. And for their sakes, return Thou on high: the Lord judgeth the people.
 	
 	9. Judge me, O Lord, according to my justice, and accordinf to my innocence on me.
 	
 	10. The wickedness of sinners shall be brought to naught; and Thou shalt direct the just: the searcher of hearts and reins is God.
 	
 	11. Just is my help from the Lord: who saveth the upright of heart.
 	
 	12. God is a just judge, strong and patient: is He angry every day?
 	
 	13. Unless ye be converted, He will brandish His sword: he hath bent His bow, and made it ready.
 	
 	14. And in it He hath prepared the instruments of death: He hath made ready His burning arrows.
 	
 	15. Behold, he hath been in labor with injustice: he hath conceived sorrow and brought forth iniquity.
 	
 	16. He hath opened a pit and dug it: and he is fallen into the hole he made.
 	
 	17. His sorrow shall be turned on his own head: and his iniquty shall come down upon his crown.
 	
 	18. I will give glory to the Lord, according to His justice: and will sing to the name of the Lord Most High.
 	
 	19. Eternal rest grant unto them, O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}

\gresetinitiallines{1}
\gregorioscore{antiphon_nequando_rapiat}
\textit{Lest at any time, like a lion, he seize upon my soul, while there is no one to rescue me, nor to save.}

%%

\subsection*{First Lesson - Job 7:16-21}

\begin{latinenglishsection}
\latinenglish{
	Parce mihi, Dómine, nihil enim sunt dies mei. Quid est homo, quia magníficas eum? aut quid appónis erga eum cor tuum? Vísitas eum dilúculo, et súbito probas illum. Usquequo non parcis mihi, nec dimíttis me ut glútiam salívam meam? Peccávi; quid fáciam tibi, O custos hóminum? quare posuísti me contrárium tibi, et factus sum mihimetípsi gravis? Cur non tollis peccátum meum, et quare non aufters iniquitátem meam? Ecce nunc in púlvere dórmiam: et, si mane me qu\ae síeris, non subsístam.
	
	{\color{red}\Rbar.} Credo quod Redémptor meus vivit, et in novíssimo die de terra surrectúrus sum. * Et in carne mea vidébo Deum Salvatórem meum.
	
	{\color{red}\Vbar.} Quem visúrus sum ego ipse et non álius, et óculi mei conspectúri sunt. * Et in carne mea vidébo Deum Salvatórem meum.
}{
	Spare me, for my days are nothing. What is man that Thou shouldst magnify him? or why dost Thou set Thy heart upon him? Thou visitest him early in the morning, and Thou provest him suddenly. How long wilt Thou not spare me, nor suffer me to swallow down my spittle? I have sinned, what shall I do to Thee, O keeper of men? why hast Thou set me opposite to Thee, and I am become burdensome to myself? Why dost Thou not remove my sin, and why dost Thou not take away mine iniquity? Behold now, I shall sleep in the dust: and if Thou seek me in the morning, I shall not be.
	
	{\color{red}\Rbar.} I believe my Redeemer liveth, and that in the last day I shall rise from the earth. * And in my flesh I shall see God my Saviour.
	
	{\color{red}\Vbar.} Whom I myself shall see, and not another, and my eyes shall behold. * And in my flesh I shall see God my Savior.
}
\end{latinenglishsection}

\subsection*{Second Lesson - Job 10:1-7}

\begin{latinenglishsection}
\latinenglish{
	Taedet ánimam meam vit\ae me\ae, dimíttam advérsum me elóquium meum, loquar in amaritúdine ánim\ae me\ae. Dicam Deo: Noli me condemnár; indica mihi, cur me ita júdices. Numquid bonum tibi vidétur, si calumniéris me, et ópprimas me opus mánuum tuárum, et consílium impiórum ádjuves? Numquid óculi cárnei tibi sunt; aut sicut videt homo, et tu vidébis? Numquid sicut dies hóminis dies tui, et anni tui sicut humána sunt témpora, ut qu\ae ras iniquitátem meam, et peccátum meum scrutéris? Et scias quia nihil ímpium fécerim, cum sit nemo qui de manu tua possit erúere.
	
	{\color{red}\Rbar.} Qui Lázarum resuscitásti a monuménto f\oe tidum: * Tu eis, Dómine, dona réquiem, et locum indulgénti\ae.
	
	{\color{red}\Vbar.} Qui ventúrus es judicáre vivos et mórtuos, et s\'{\ae}culum per ignem: * Tu eis, Dómine, dona réquiem, et locum indulgénti\ae.
}{
	My soul is weary of life, I will let go my speech against myself, I will speak in the bitterness of my soul, I will say to God: Do not condemn me; tell me why Thou judgest me so. Doth it seem good to Thee that Thou shouldst crush me and oppress me, the work of Thy own hands, and help the counsel of the wicked? Hast Thou eyes of flesh, or seest Thou as man seeth? Are Thy days as the days of man, and are Thy years as the times of men, Thou shouldst inquire after my iniquity, and search after my sin? And Thou knowest that I have done no wickerd thing, whereas there is no man that can deliver out of Thy hand.
	
	{\color{red}\Rbar.} Thou who didst raise Lazarus from the grave: * Thou, O Lord, give them rest, and a place of pardon.
	
	{\color{red}\Vbar.} Thou who art to come to judge the living and the dead, and the world by fire: * Thou, O Lord, give them rest, and a place of pardon.
}
\end{latinenglishsection}

\subsection*{Third Lesson - Job 10:8-12}

\begin{latinenglishsection}
\latinenglish{
	Manus tu\ae fec\'{e}runt me, et plasmav\'{e}runt me totum in circ\'{u}itu: et sic rep\'{e}nte pr\ae c\'{i}pitas me? Mem\'{e}nto qu\ae so quod sicut lutum f\'{e}ceris me, et in p\'{u}lverem red\'{u}ces me. Nonne sicut lac muls\'{i}sti me, et sicut c\'{a}seum me coagul\'{a}sto? Pelle et c\'{a}rnibus vest\'{i}sti me; \'{o}ssibus et nervis compeg\'{i}sti me. Vitam et miseric\'{o}rdiam tribu\'{i}sti mihi, et visit\'{a}tio tua custod\'{i}vit sp\'{i}ritum meum.
	
	{\color{red}\Rbar.} D\'{o}mine, quando v\'{e}neris judic\'{a}re terram, ubi me absc\'{o}ndam a vultu ir\ae tu\ae? * Quia pecc\'{a}vi nimis in vita mea.
	
	{\color{red}\Vbar.} Comm\'{i}ssa mea pav\'{e}sco, et ante te erub\'{e}sco: dum v\'{e}neris judic\'{a}re, noli me condemn\'{a}re. * Quia pecc\'{a}vi nimis in vita mea.
	
	{\color{red}\Rbar.} R\'{e}quiem \ae t\'{e}rnam dona eis, D\'{o}mine: et lux perp\'{e}tua l\'{u}ceat eis. * Quia pecc\'{a}vi nimis in vita mea.
}{
	Thy hands have made me, and fashioned me wholly round about: and dost Thou thus cast me down headlong on a sudden? Remember, I beseech Thee, that Thou hast made me as the clay, and Thou wilt bring me into dust again. Hast Thou not pressed me out as milk, and curdled me like cheese? Thou hast clothed me with skin and flesh; Thou hast put me together with bones and sinews. Thou hast granted me life and mercy, and Thy visitation hath preserved my spirit.	
	
	{\color{red}\Rbar.} O Lord, when Thou wilt come to judge the earth, where shall I hide myself from the face of Thy wrath? * For I have sinned exceedingly in my life.
	
	{\color{red}\Vbar.} I dread my misdeeds, and blush before Thee; do not condemn me when Thou comest me when Thou comest to judge. * For I have sinned exceedingly in my life.
	
	{\color{red}\Rbar.} Eternal rest grant unto them, O Lord, and let perpetual light shine upon them. * For I have sinned exceedingly in my life.
}
\end{latinenglishsection}

{\color{red}\textit{If only the First Nocturn is said, then Lauds is immediately recited (pg. ). Otherwise, if Lauds is to be omitted, then continue with the Pater Noster and prayers beginning on pg. .}}
 
\vfill\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

 \section*{Second Nocturn}
 \begin{center}
 \textit{To be said on Tuesdays and Fridays}
 \end{center}
 
  \textit{He hath set me in a place of pasture.}
 \gresetinitiallines{1}
 \gregorioscore{antiphon_in_loco_pascuae}
 
 \subsection*{Psalm 22}
 
 \gresetinitiallines{1}
 \gregorioscore{intonation_psalm_22}
 
 \begin{latinenglishsection} 
 \latinenglish{
 	2. Super aquam refectiónis edu\textbf{cá}vit me:~* ánimam me\textit{am} \textit{con}\textbf{vér}tit.

	3. Dedúxit me super sémitas jus\textbf{tí}tiæ:~* propter \textit{no}\textit{men} \textbf{su}um.
	
	4. Nam, etsi ambulávero in médio umbræ mortis, non timébo \textbf{ma}la:~* quóni\textit{am} \textit{tu} \textbf{me}cum es.
	
	5. Virga tua, et báculus \textbf{tu}us:~* ipsa me \textit{con}\textit{so}\textbf{lá}ta sunt.
	
	6. Parásti in conspéctu meo \textbf{men}sam,~* advérsus eos, qui \textit{trí}\textit{bu}\textbf{lant} me.
	
	7. Impinguásti in óleo caput \textbf{me}um:~* et calix meus inébrians \textit{quam} \textit{præ}\textbf{clá}rus est!
	
	8. Et misericórdia tua subse\textbf{qué}tur me~* ómnibus diébus \textit{vi}\textit{tæ} \textbf{me}æ.
	
	9. Et ut inhábitem in domo \textbf{Dó}mini,~* in longitúdi\textit{nem} \textit{di}\textbf{é}rum.
	
	10. Requiem æternam dona eis, \textbf{Do}mine~* Et lux perpetua lu\textit{ceat} \textbf{e}is.
 }{
 	1. The Lord ruleth me: and I shall want nothing: He hath set me in a place of pasture.
 	
 	2. He hath brought me up, on the water of refreshment: He hath converted my soul. 
 	
 	3. He hath led me on the paths of justice, for his own name's sake.
 	
 	4. For though I should walk in the midst of the shadow of death, I will fear no evils, for thou art with me.
 	
 	5. Thy rod and thy staff, they have comforted me.
 	
 	6. Thou hast prepared a table before me against them that afflict me. 
 	
 	7. Thou hast anointed my head with oil; and my chalice which inebriateth me, how goodly is it!
 	
 	8. And thy mercy will follow me all the days of my life.
 	
 	9. And that I may dwell in the house of the Lord unto length of days.
 	
 	10. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}
 
\gresetinitiallines{1}
\gregorioscore{antiphon_in_loco_pascuae}
\textit{He hath set me in a place of pasture.}

%%
 
\vfill\pagebreak %\hspace{2cm}
 
\textit{The sins of my youth and my offences, do not remember, O Lord.}
\gresetinitiallines{1}
\gregorioscore{antiphon_delicta_juventutis}
 
\subsection*{Psalm 24}
 
\gresetinitiallines{1}
\gregorioscore{intonation_psalm_24}

 \begin{latinenglishsection} 
 \latinenglish{
 	2. Neque irrídeant me inimíci \textbf{me}i:~* étenim univérsi, qui sústinent te, non \textit{con}\textit{fun}\textbf{dén}tur.

	3. Confundántur omnes iníqua a\textbf{gén}tes~* \textit{su}\textit{per}\textbf{vá}cue.
	
	4. Vias tuas, Dómine, demónstra \textbf{mi}hi~* et sémitas tuas \textit{é}\textit{do}\textbf{ce} me.
	
	5. Dírige me in veritáte tua, et \textbf{do}ce me:~* quia tu es, Deus, Salvátor meus, et te sustínui \textit{to}\textit{ta} \textbf{di}e.
	
	6. Reminíscere miseratiónum tuárum, \textbf{Dó}mine,~* et misericordiárum tuárum, quæ a \textit{sǽ}\textit{cu}\textbf{lo} sunt.
	
	7. Delícta juventútis \textbf{me}æ,~* et ignorántias meas \textit{ne} \textit{me}\textbf{mí}neris.
	
	8. Secúndum misericórdiam tuam meménto \textbf{me}i tu,~* propter bonitátem \textit{tu}\textit{am}, \textbf{Dó}mine.
	
	9. Dulcis et rectus \textbf{Dó}minus:~* propter hoc legem dabit delinquénti\textit{bus} \textit{in} \textbf{vi}a.
	
	10. Díriget mansuétos in ju\textbf{dí}cio:~* docébit mites \textit{vi}\textit{as} \textbf{su}as.
	
	11. Univérsæ viæ Dómini, misericórdia et \textbf{vé}ritas,~* requiréntibus testaméntum ejus et testimó\textit{ni}\textit{a} \textbf{e}jus.
	
	12. Propter nomen tuum, Dómine, propitiáberis peccáto \textbf{me}o:~* mul\textit{tum} \textit{est} \textbf{e}nim.
	
	13. Quis est homo qui timet \textbf{Dó}minum?~* legem státuit ei in via, \textit{quam} \textit{e}\textbf{lé}git.
	
	14. Anima ejus in bonis demo\textbf{rá}bitur:~* et semen ejus heredi\textit{tá}\textit{bit} \textbf{ter}ram.
	
	15. Firmaméntum est Dóminus timéntibus \textbf{e}um:~* et testaméntum ipsíus ut manifes\textit{té}\textit{tur} \textbf{il}lis.
	
	16. Oculi mei semper ad \textbf{Dó}minum:~* quóniam ipse evéllet de láqueo \textit{pe}\textit{des} \textbf{me}os.
	
	17. Réspice in me, et miserére \textbf{me}i:~* quia únicus et pau\textit{per} \textit{sum} \textbf{e}go.
	
	18. Tribulatiónes cordis mei multipli\textbf{cá}tæ sunt:~* de necessitátibus meis \textit{é}\textit{ru}\textbf{e} me.
	
	19. Vide humilitátem meam, et labórem \textbf{me}um:~* et dimítte univérsa de\textit{líc}\textit{ta} \textbf{me}a.
	
	20. Réspice inimícos meos quóniam multipli\textbf{cá}ti sunt:~* et ódio iní\textit{quo} \textit{o}\textbf{dé}runt me.
	
	21. Custódi ánimam meam, et éru\textbf{e} me:~* non erubéscam quóniam spe\textit{rá}\textit{vi} \textbf{in} te.
	
	22. Innocéntes et recti adhæsérunt \textbf{mi}hi:~* quia sus\textit{tí}\textit{nu}\textbf{i} te.
	
	23. Líbera, Deus, \textbf{Is}raël:~* ex ómnibus tribulatió\textit{ni}\textit{bus} \textbf{su}is.
	
	24. Requiem æternam dona eis, \textbf{Do}mine~* Et lux perpetua lu\textit{ceat} \textbf{e}is.
 }{
 	1. To thee, O Lord, have I lifted up my soul. In thee, O my God, I put my trust; let me not be ashamed.
 	
 	2. Neither let my enemies laugh at me: for none of them that wait on thee shall be confounded.
 	
 	3. Let all them be confounded that act unjust things without cause. 
 	
 	4. Shew, O Lord, thy ways to me, and teach me thy paths.
 	
 	5. Direct me in thy truth, and teach me; for thou art God my Saviour; and on thee have I waited all the day long.
 	
 	6. Remember, O Lord, thy bowels of compassion; and thy mercies that are from the beginning of the world.
 	
 	7. The sins of my youth and my ignorances do not remember. 
 	
 	8. According to thy mercy remember thou me: for thy goodness' sake, O Lord.
 	
 	9. The Lord is sweet and righteous: therefore he will give a law to sinners in the way.
 	
 	10. He will guide the mild in judgment: he will teach the meek his ways.
 	
 	11. All the ways of the Lord are mercy and truth, to them that seek after his covenant and his testimonies.
 	
 	12. For thy name's sake, O Lord, thou wilt pardon my sin: for it is great.
 	
 	13. Who is the man that feareth the Lord? He hath appointed him a law in the way he hath chosen.
 	
 	14. His soul shall dwell in good things: and his seed shall inherit the land.
 	
 	15. The Lord is a firmament to them that fear him: and his covenant shall be made manifest to them.
 	
 	16. My eyes are ever towards the Lord: for he shall pluck my feet out of the snare.
 	
 	17. Look thou upon me, and have mercy on me; for I am alone and poor.
 	
 	18. The troubles of my heart are multiplied: deliver me from my necessities.
 	
 	19. See my abjection and my labour; and forgive me all my sins.
 	
 	20. Consider my enemies for they are multiplied, and have hated me with an unjust hatred.
 	
 	21. Keep thou my soul, and deliver me: I shall not be ashamed, for I have hoped in thee.
 	
 	22. The innocent and the upright have adhered to me: because I have waited on thee.
 	
 	23. Deliver Israel, O God, from all his tribulations.
 	
 	24. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}

\gresetinitiallines{1}
\gregorioscore{antiphon_delicta_juventutis}
\textit{The sins of my youth and my offences, do not remember, O Lord.}

%%

\hspace{2cm}

\textit{I believe that I shall see the good things of the Lord in the land of the living.}
\gresetinitiallines{1}
\gregorioscore{antiphon_credo_videre}
 
\subsection*{Psalm 26}
 
\gresetinitiallines{1}
\gregorioscore{intonation_psalm_26}

 \begin{latinenglishsection} 
 \latinenglish{
 	2. Dóminus protéctor \textit{vi}\textit{tæ} \textbf{me}æ,~* a \textit{quo} \textit{tre}\textit{pi}\textbf{dá}bo?

	3. Dum apprópiant super \textit{me} \textit{no}\textbf{cén}tes,~* ut e\textit{dant} \textit{car}\textit{nes} \textbf{me}as:
	
	4. Qui tríbulant me ini\textit{mí}\textit{ci} \textbf{me}i,~* ipsi infirmáti sunt \textit{et} \textit{ce}\textit{ci}\textbf{dé}runt.
	
	5. Si consístant advér\textit{sum} \textit{me} \textbf{cas}tra,~* non ti\textit{mé}\textit{bit} \textit{cor} \textbf{me}um.
	
	6. Si exsúrgat advér\textit{sum} \textit{me} \textbf{prǽ}lium,~* in hoc \textit{e}\textit{go} \textit{spe}\textbf{rá}bo.
	
	7. Unam pétii a Dómino, \textit{hanc} \textit{re}\textbf{quí}ram,~* ut inhábitem in domo Dómini ómnibus dié\textit{bus} \textit{vi}\textit{tæ} \textbf{me}æ:
	
	8. Ut vídeam volup\textit{tá}\textit{tem} \textbf{Dó}mini,~* et vísi\textit{tem} \textit{tem}\textit{plum} \textbf{e}jus.
	
	9. Quóniam abscóndit me in taberná\textit{cu}\textit{lo} \textbf{su}o:~* in die malórum protéxit me in abscóndito taber\textit{ná}\textit{cu}\textit{li} \textbf{su}i.
	
	10. In petra \textit{ex}\textit{al}\textbf{tá}vit me:~* et nunc exaltávit caput meum super in\textit{i}\textit{mí}\textit{cos} \textbf{me}os.
	
	11. Circuívi et immolávi in tabernáculo ejus hóstiam vocife\textit{ra}\textit{ti}\textbf{ó}nis:~* cantábo et psal\textit{mum} \textit{di}\textit{cam} \textbf{Dó}\textbf{mi}no.
	
	12. Exáudi, Dómine, vocem meam, qua cla\textit{má}\textit{vi} \textbf{ad} te:~* miserére me\textit{i}, \textit{et} \textit{ex}\textbf{áu}\textbf{di} me.
	
	13. Tibi dixit cor meum, exquisívit te fá\textit{ci}\textit{es} \textbf{me}a:~* fáciem tuam, Dó\textit{mi}\textit{ne}, \textit{re}\textbf{quí}ram.
	
	14. Ne avértas fáciem \textit{tu}\textit{am} \textbf{a} me,~* ne declínes in ira \textit{a} \textit{ser}\textit{vo} \textbf{tu}o.
	
	15. Adjútor \textit{me}\textit{us} \textbf{es}to:~* ne derelínquas me, neque despícias me, Deus, sa\textit{lu}\textit{tá}\textit{ris} \textbf{me}us.
	
	16. Quóniam pater meus, et mater mea de\textit{re}\textit{li}\textbf{qué}runt me:~* Dóminus \textit{au}\textit{tem} \textit{as}\textbf{súmp}\textbf{sit} me.
	
	17. Legem pone mihi, Dómine, in \textit{vi}\textit{a} \textbf{tu}a:~* et dírige me in sémitam rectam propter in\textit{i}\textit{mí}\textit{cos} \textbf{me}os.
	
	18. Ne tradíderis me in ánimas tribu\textit{lán}\textit{ti}\textbf{um} me:~* quóniam insurrexérunt in me testes iníqui, et mentíta est in\textit{í}\textit{qui}\textit{tas} \textbf{si}bi.
	
	19. Credo vidére \textit{bo}\textit{na} \textbf{Dó}mini~* in \textit{ter}\textit{ra} \textit{vi}\textbf{vén}\textbf{ti}um.
	
	20. Exspécta Dóminum, virí\textit{li}\textit{ter} \textbf{a}ge:~* et confortétur cor tuum, et \textit{sús}\textit{ti}\textit{ne} \textbf{Dó}\textbf{mi}num.
	
	21. Requiem æternam dona \textit{eis}, \textbf{Do}mine~* Et lux perpetua lu\textit{ceat} \textbf{e}is.
 }{
 	1. The Lord is my light and my salvation, whom shall I fear?
 	
 	2. The Lord is the protector of my life: of whom shall I be afraid?
 	
 	3. Whilst the wicked draw near against me, to eat my flesh.
 	
 	4. My enemies that trouble me, have themselves been weakened, and have fallen.
 	
 	5. If armies in camp should stand together against me, my heart shall not fear.
 	
 	6. If a battle should rise up against me, in this will I be confident.
 	
 	7. One thing I have asked of the Lord, this will I seek after; that I may dwell in the house of the Lord all the days of my life. 
 	
 	8. That I may see the delight of the Lord, and may visit his temple.
 	
 	9. For he hath hidden me in his tabernacle; in the day of evils, he hath protected me in the secret place of his tabernacle.
 	
 	10. He hath exalted me upon a rock: and now he hath lifted up my head above my enemies.
 	
 	11. I have gone round, and have offered up in his tabernacle a sacrifice of jubilation: I will sing, and recite a psalm to the Lord.
 	
 	12. Hear, O Lord, my voice, with which I have cried to thee: have mercy on me and hear me.
 	
 	13. My heart hath said to thee: My face hath sought thee: thy face, O Lord, will I still seek.
 	
 	14. Turn not away thy face from me; decline not in thy wrath from thy servant.
 	
 	15. Be thou my helper, forsake me not; do not thou despise me, O God my Saviour.
 	
 	16. For my father and my mother have left me: but the Lord hath taken me up.
 	
 	17. Set me, O Lord, a law in thy way, and guide me in the right path, because of my enemies.
 	
 	18. Deliver me not over to the will of them that trouble me; for unjust witnesses have risen up against me; and iniquity hath lied to itself.
 	
 	19. I believe to see the good things of the Lord in the land of the living.
 	
 	20. Expect the Lord, do manfully, and let thy heart take courage, and wait thou for the Lord.
 	
 	21. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}

\gresetinitiallines{1}
\gregorioscore{antiphon_credo_videre}
 \textit{I believe that I shall see the good things of the Lord in the land of the living.}

 %%
 
 \subsection*{Fourth Lesson - Job 13:22-28}
 
 \begin{latinenglishsection}
\latinenglish{
	Responde mihi: Quantas h\'{a}beo iniquit\'{a}tes et pecc\'{a}ta' sc\'{e}lera mea et del\'{i}cta ost\'{e}nde mihi. Cur f\'{a}ciem tuam absc\'{o}ndis, et arbitr\'{a}ris me inim\'{i}cum tuum? Contra f\'{o}lium, quod vento r\'{a}pitur, ost\'{e}ndis pot\'{e}ntiam tuam, et st\'{i}pulam siccam pers\'{e}queris. Scribis enim contra me amarit\'{u}dines, et cons\'{u}mere me vis pecc\'{a}tis adolesc\'{e}nti\ae me\ae. Posu\'{i}sti in nervo pedem meum, et observ\'{a}sti omnes s\'{e}mitas meas, et vert\'{i}gia pedum me\'{o}rum consider\'{a}sti; qui quasi putr\'{e}do consum\'{e}ndus sum, et quasi vestim\'{e}ntum, quod com\'{e}ditur a t\'{i}nea.
	
	{\color{red}\Rbar.} Mem\'{e}nto mei, Deus, quia ventus est vita mea, * nec asp\'{i}ciat me visus h\'{o}minis.
	
	{\color{red}\Vbar.} De prof\'{u}ndis clam\'{a}vi ad te, D\'{o}mine: D\'{o}mine, ex\'{a}udi vocem meam. * Nec asp\'{i}ciat me visus h\'{o}minis.
}{
	I will speak, and do Thou answer me. How many are my iniquities and sins? make me know my crimes and offences. Why hidest Thou Thy face, and thinkest me Thy enemy? Against a leaf that is carried away with the wind, Thou showest Thy power; and Thou pursuest dry stubble. For Thou writest bitter things against me, and consumest me for the sins of my youth. Thou hast put my feet in the stocks, and Thou observest all my paths, and considerest the steos if ny feet: who am to be consumed as rottenness, and as a garment that is moth-eaten.
	
	{\color{red}\Rbar.} Remember me, O God, because my life is but wind, * nor may the sight of man behold me.
	
	{\color{red}\Vbar.} From the depths I have cried to Thee, O Lord; Lord, hear my voice. * Nor may the sight of man behold me.
}
\end{latinenglishsection}
\vfill\pagebreak

\subsection*{Fifth Lesson - Job 14:1-6}

 \begin{latinenglishsection}
\latinenglish{
	Homo natus de mul\'{i}ere, brevi vivens t\'{e}mpore, repl\'{e}tur multis mis\'{e}riis. Qui quasi flos egr\'{e}dietur et cont\'{e}ritur, etfugit velut umbra, et numquam in e\'{o}dem statu p\'{e}rmanet. Et dignum ducis super hujusc\'{e}modi aper\'{i}re \'{o}culos tuos, et add\'{u}cere eum tecu in jud\'{i}cium? Quis potest f\'{a}cere mundum de imm\'{u}ndo conc\'{e}ptum s\'{e}mine? Nonne tu quis solus es? Breves dies h\'{o}minis sunt, numerus m\'{e}nsium ejus apud te est: Constitu\'{i}sti t\'{e}rminos ejus, qui pr\ae ter\'{i}ri non p\'{o}terunt. Rec\'{e}de p\'{a}ululum ab eo, ut qui\'{e}scat, donec opt\'{a}ta v\'{e}niat, sicut mercen\'{a}rii, dies ejus.
	
	{\color{red}\Rbar.} Heu mihi, D\'{o}mine, quia pecc\'{a}vi nimis in vita mea: Quid f\'{a}ciam, miser? ubi f\'{u}giam, nisi ad te, Deus meus? * Miser\'{e}re mei, dum v\'{e}neris in nov\'{i}ssimo die.
	
	{\color{red}\Vbar.} Anima mea turb\'{a}ta est valde, sed tu, D\'{o}mine, succ\'{u}rre ei. * Miser\'{e}re mei, dum v\'{e}neris in nov\'{i}ssimo die.
}{
	Man born of woman, living for a short time, is filled with many miseries. He cometh forth like a flower, and is destroyed, and fleeth as a shadow, and never continueth in the same state. And dost Thou think it meet to open Thy eyes upon such a one, and to bring him into judgement with Thee? Who can make him clean that is conceived of unclean seed? Is it not Thou only? The days of man are short, and the number of his months is with Thee: Thou hast appointed his bounds, which cannot be passed. Depart a little from him, that he may rest, until his wished-for day come, as that of the hireling.
	
	{\color{red}\Rbar.} Woe is me, O Lord, because I have sinned exceedingly in my life; O wretch, what shall I do, whither shall I fly unless to Thee, my God? * Have mercy on me, when Thou comest at the latter day.
	
	{\color{red}\Vbar.} My soul is greatly troubled; but Thou, O Lord, succor it. * Have mercy on me, when Thou comest at the latter day.
}
\end{latinenglishsection}

\subsection*{Sixth Lesson - Job 14:13-16}

 \begin{latinenglishsection}
\latinenglish{
	Quis mihi hoc tr\'{i}buat, ut in inf\'{e}rno pr\'{o}tegas me et absc\'{o}ndas me, donec pertr\'{a}nseat furor tuus, et const\'{i}tuas mihi tempus in quo record\'{e}ris mei? Purt\'{a}sne m\'{o}rtuus homo rursum vivat? Cunctis di\'{e}bus quibus nunc m\'{i}lito, exp\'{e}cto donec v'{e}niat immut\'{a}tio mea. Voc\'{a}bus me, et ego respond\'{e}bo tibi; \'{o}peri m\'{a}nuum tu\'{a}rum p\'{o}rriges d\'{e}xteram. Tu quidem gressus meos dinumer\'{a}sti, sed parce pecc\'{a}tis meis.
	
	{\color{red}\Rbar.} Ne record\'{e}ris pecc\'{a}ta mea, D\'{o}mine: * Dum v\'{e}neris judic\'{a}re s\'{\ae}culum per ignem.
	
	{\color{red}\Vbar.} D\'{i}rige, D\'{o}mine Deus meus, in consp\'{e}ctu tuo viam meam. * Dum v\'{e}neris judicare s\'{\ae}culum per ignem.
	
	{\color{red}\Vbar.} R\'{e}quiem \ae t\'{e}rnam dona eis, D\'{o}mine: et lux perp\'{e}tua l\'{u}ceat eis. * Dum v\'{e}neris judic\'{a}re s\'{\ae}culum per ignem.
}{
	Who will grant me this, that Thou mayest protect me in hell, and hide me till Thy wrath pass, and appoint me a time when Thou wilt remember me? Shall man that is dead, thinkest Thou, live again? All the days in which I am now in warfare, I expect until my change come. Thou wilt call me, and I will anwer Thee: to the work of Thy hands Thou wilt reach out Thy right hand. Thou indeed hast numbered my steps, but spare my sins.
	
	{\color{red}\Rbar.} Remember not my sins, O Lord: * When Thou wilt come to judge the world by fire.
	
	{\color{red}\Vbar.} Direct, O Lord my God, my way in Thy sight. * When Thou wilt come to judge the world by fire.
	
	{\color{red}\Vbar.} Eternal rest grant unto them, O Lord, and let perptual light shine upon them. * When Thou wilt come to judge the world by fire.
}
\end{latinenglishsection}

{\color{red}\textit{If only the Second Nocturn is said, then Lauds is immediately recited (pg. ). Otherwise, if Lauds is to be omitted, then continue with the Pater Noster and prayers beginning on pg. .}}
 
 \vfill\pagebreak
 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section*{Third Nocturn}
 \begin{center}
 \textit{To be said on Wednesdays and Saturdays.}
 \end{center}
 
\textit{May it please Thee, O Lord, to deliver me: Lord, have regard to help me.}
\gresetinitiallines{1}
\gregorioscore{antiphon_complaceat_tibi}
 
\subsection*{Psalm 39}
 
 \gresetinitiallines{1}
 \gregorioscore{intonation_psalm_39}
 
 \begin{latinenglishsection} 
 \latinenglish{
 	2. Et exaudívit preces \textbf{me}as:~* et edúxit me de lacu misériæ, et de lu\textit{to} \textbf{fæ}cis.

	3. Et státuit super petram pedes \textbf{me}os:~* et diréxit gres\textit{sus} \textbf{me}os.
	
	4. Et immísit in os meum cánticum \textbf{no}vum~* carmen De\textit{o} \textbf{nos}tro.
	
	5. Vidébunt multi, et ti\textbf{mé}bunt:~* et sperábunt \textit{in} \textbf{Dó}mino.
	
	6. Beátus vir, cujus est nomen Dómini spes \textbf{e}jus~* et non respéxit in vanitátes et insáni\textit{as} \textbf{fal}sas.
	
	7. Multa fecísti tu, Dómine, Deus meus, mirabília \textbf{tu}a:~* et cogitatiónibus tuis non est qui símilis \textit{sit} \textbf{ti}bi.
	
	8. Annuntiávi et lo\textbf{cú}tus sum:~* multiplicáti sunt su\textit{per} \textbf{nú}merum.
	
	9. Sacrifícium et oblatiónem nolu\textbf{ís}ti:~* aures autem perfecís\textit{ti} \textbf{mi}hi.
	
	10. Holocáustum et pro peccáto non postu\textbf{lás}ti:~* tunc dixi: Ec\textit{ce} \textbf{vé}nio.
	
	11. In cápite libri scriptum est de me ut fácerem voluntátem \textbf{tu}am:~* Deus meus, vólui, et legem tuam in médio cor\textit{dis} \textbf{me}i.
	
	12. Annuntiávi justítiam tuam in ecclésia \textbf{ma}gna,~* ecce lábia mea non prohibébo: Dómine, \textit{tu} \textbf{scis}ti.
	
	13. Justítiam tuam non abscóndi in corde \textbf{me}o:~* veritátem tuam et salutáre tu\textit{um} \textbf{di}xi.
	
	14. Non abscóndi misericórdiam tuam et veritátem \textbf{tu}am~* a concíli\textit{o} \textbf{mul}to.
	
	15. Tu autem, Dómine, ne longe fácias miseratiónes tuas \textbf{a} me:~* misericórdia tua et véritas tua semper su\textit{sce}\textbf{pé}runt me.
	
	16. Quóniam circumdedérunt me mala, quorum non est \textbf{nú}merus:~* comprehendérunt me iniquitátes meæ, et non pótui ut \textit{vi}\textbf{dé}rem.
	
	17. Multiplicátæ sunt super capíllos cápitis \textbf{me}i:~* et cor meum de\textit{re}\textbf{lí}quit me.
	
	18. Compláceat tibi, Dómine, ut éru\textbf{as} me:~* Dómine, ad adjuvándum \textit{me} \textbf{ré}spice.
	
	19. Confundántur et revereántur simul, qui quærunt ánimam \textbf{me}am,~* ut áufe\textit{rant} \textbf{e}am.
	
	20. Convertántur retrórsum et revere\textbf{án}tur:~* qui volunt mi\textit{hi} \textbf{ma}la.
	
	21. Ferant conféstim confusiónem \textbf{su}am:~* qui dicunt mihi: Eu\textit{ge}, \textbf{eu}ge.
	
	22. Exsúltent et læténtur super te omnes quæ\textbf{rén}tes te:~* et dicant semper: Magnificétur Dóminus: qui díligunt salutá\textit{re} \textbf{tu}um.
	
	23. Ego autem mendícus sum, et \textbf{pau}per:~* Dóminus sollícitus \textit{est} \textbf{me}i.
	
	24. Adjútor meus et protéctor meus \textbf{tu} es:~* Deus meus, ne \textit{tar}\textbf{dá}veris.
	
	25. Requiem æternam dona eis, \textbf{Do}mine~* Et lux perpetua lucea\textit{at} \textbf{e}is.
 }{
 	1. With expectation I have waited for the Lord, and he was attentive to me.
 	
 	2. And he heard my prayers, and brought me out of the pit of misery and the mire of dregs.
 	
 	3. And he set my feet upon a rock, and directed my steps.
 	
 	4. And he put a new canticle into my mouth, a song to our God. 
 	
 	5. Many shall see, and shall fear: and they shall hope in the Lord.
 	
 	6. Blessed is the man whose trust is in the name of the Lord; and who hath not had regard to vanities, and lying follies.
 	
 	7. Thou hast multiplied thy wonderful works, O Lord my God: and in thy thoughts there is no one like to thee.
 	
 	8. I have declared and I have spoken they are multiplied above number.
 	
 	9. Sacrifice and oblation thou didst not desire; but thou hast pierced ears for me.
 	
 	10. Burnt offering and sin offering thou didst not require: Then said I, Behold I come.
 	
 	11. In the head of the book it is written of me that I should do thy will: O my God, I have desired it, and thy law in the midst of my heart.
 	
 	12. I have declared thy justice in a great church, lo, I will not restrain my lips: O Lord, thou knowest it.
 	
 	13. I have not hid thy justice within my heart: I have declared thy truth and thy salvation.
 	
 	14. I have not concealed thy mercy and thy truth from a great council.
 	
 	15. Withhold not thou, O Lord, thy tender mercies from me: thy mercy and thy truth have always upheld me.
 	
 	16. For evils without number have surrounded me; my iniquities have overtaken me, and I was not able to see.
 	
 	17. They are multiplied above the hairs of my head: and my heart hath forsaken me.
 	
 	18. Be pleased, O Lord, to deliver me, look down, O Lord, to help me.
 	
 	19. Let them be confounded and ashamed together, that seek after my soul to take it away.
 	
 	20. Let them be turned backward and be ashamed that desire evils to me.
 	
 	21. Let them immediately bear their confusion, that say to me: 'T is well, 't is well.
 	
 	22.  Let all that seek thee rejoice and be glad in thee: and let such as love thy salvation say always: The Lord be magnified.
 	
 	23. But I am a beggar and poor: the Lord is careful for me.
 	
 	24. Thou art my helper and my protector: O my God, be not slack.
 	
 	25. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}
 
 \gresetinitiallines{1}
 \gregorioscore{antiphon_complaceat_tibi}
 \textit{May it please Thee, O Lord, to deliver me: Lord, have regard to help me.}
 
 %%
 
\hspace{2cm}
 
\textit{Heal my soul, O Lord, for I have sinned against Thee.}
\gresetinitiallines{1}
\gregorioscore{antiphon_sana_domine}
 
\subsection*{Psalm 40}
 
 \gresetinitiallines{1}
 \gregorioscore{intonation_psalm_40}
 
 \begin{latinenglishsection} 
 \latinenglish{
 	2. Dóminus consérvet eum, et vivíficet eum, et beátum fáciat eum in \textbf{ter}ra:~* et non tradat eum in ánimam inimicó\textit{rum} \textbf{e}jus.

	3. Dóminus opem ferat illi super lectum dolóris \textbf{e}jus:~* univérsum stratum ejus versásti in infirmitá\textit{te} \textbf{e}jus.
	
	4. Ego dixi: Dómine, miserére \textbf{me}i:~* sana ánimam meam, quia peccá\textit{vi} \textbf{ti}bi.
	
	5. Inimíci mei dixérunt mala \textbf{mi}hi:~* Quando moriétur, et períbit no\textit{men} \textbf{e}jus?
	
	6. Et si ingrediebátur ut vidéret, vana loque\textbf{bá}tur:~* cor ejus congregávit iniquitá\textit{tem} \textbf{si}bi.
	
	7. Egrediebátur \textbf{fo}ras,~* et loquebátur in \textit{id}\textbf{íp}sum.
	
	8. Advérsum me susurrábant omnes inimíci \textbf{me}i:~* advérsum me cogitábant ma\textit{la} \textbf{mi}hi.
	
	9. Verbum iníquum constituérunt ad\textbf{vér}sum me:~* Numquid qui dormit non adjíciet ut \textit{re}\textbf{súr}gat?
	
	10. Etenim homo pacis meæ, in quo spe\textbf{rá}vi:~* qui edébat panes meos, magnificávit super me supplanta\textit{ti}\textbf{ó}nem.
	
	11. Tu autem, Dómine, miserére mei, et resúsci\textbf{ta} me:~* et retríbu\textit{am} \textbf{e}is.
	
	12. In hoc cognóvi quóniam volu\textbf{ís}ti me:~* quóniam non gaudébit inimícus me\textit{us} \textbf{su}per me.
	
	13. Me autem propter innocéntiam susce\textbf{pís}ti:~* et confirmásti me in conspéctu tuo in \textit{æ}\textbf{tér}num.
	
	14. Benedíctus Dóminus, Deus Israël, a sǽculo et usque in \textbf{sǽ}culum:~* fi\textit{at}, \textbf{fi}at.
	
	15. Requiem æternam dona eis, \textbf{Do}mine~* Et lux perpetua lucea\textit{at} \textbf{e}is.
 }{
 	1. Blessed is he that understandeth concerning the needy and the poor: the Lord will deliver him in the evil day.
 	
 	2. The Lord preserve him and give him life, and make him blessed upon the earth: and deliver him not up to the will of his enemies.
 	
 	3. The Lord help him on his bed of sorrow: thou hast turned all his couch in his sickness.

	4. I said: O Lord, be thou merciful to me: heal my soul, for I have sinned against thee.
	
	5. My enemies have spoken evils against me: when shall he die and his name perish?
	
	6. And if he came in to see me, he spoke vain things: his heart gathered together iniquity to itself.
	
	7. He went out, and spoke to the same purpose.
	
	8. All my enemies whispered together against me: they devised evils to me.
	
	9. They determined against me an unjust word: shall he that sleepeth rise again no more?
	
	10. For even the man of peace, in whom I trusted, who ate my bread, hath greatly supplanted me.
	
	11. But thou, O Lord, have mercy on me, and raise me up again: and I will requite them.
	
	12. By this I know, that thou hast had a good will for me: because my enemy shall not rejoice over me.
	
	13. But thou hast upheld me by reason of my innocence: and hast established me in thy sight for ever.
	
	14. Blessed by the Lord the God of Israel from eternity to eternity. So be it. So be it.
 	
 	15. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}
 
\gresetinitiallines{1}
\gregorioscore{antiphon_sana_domine}
\textit{Heal my soul, O Lord, for I have sinned against Thee.}

%%

\hspace{2cm}

\textit{My soul has thirsted after the living God: when shall I come and appear before the face of the Lord?}
\gresetinitiallines{1}
\gregorioscore{antiphon_sitivit_anima}
 
\subsection*{Psalm 41}
 
 \gresetinitiallines{1}
 \gregorioscore{intonation_psalm_41}
 
 \begin{latinenglishsection} 
 \latinenglish{
 	2. Sitívit ánima mea ad Deum \textit{for}\textit{tem} \textbf{vi}vum:~* quando véniam, et apparébo ante \textit{fá}\textit{ci}\textit{em} \textbf{De}i?

	3. Fuérunt mihi lácrimæ meæ panes di\textit{e} \textit{ac} \textbf{noc}te:~* dum dícitur mihi quotídie: Ubi \textit{est} \textit{De}\textit{us} \textbf{tu}us?
	
	4. Hæc recordátus sum, et effúdi in me á\textit{ni}\textit{mam} \textbf{me}am:~* quóniam transíbo in locum tabernáculi admirábilis, usque \textit{ad} \textit{do}\textit{mum} \textbf{De}i.
	
	5. In voce exsultatiónis, et con\textit{fes}\textit{si}\textbf{ó}nis:~* so\textit{nus} \textit{e}\textit{pu}\textbf{lán}tis.
	
	6. Quare tristis es, á\textit{ni}\textit{ma} \textbf{me}a?~* et \textit{qua}\textit{re} \textit{con}\textbf{túr}\textbf{bas} me?
	
	7. Spera in Deo, quóniam adhuc confi\textit{té}\textit{bor} \textbf{il}li:~* salutáre vultus mei, \textit{et} \textit{De}\textit{us} \textbf{me}us.
	
	8. Ad meípsum ánima mea \textit{con}\textit{tur}\textbf{bá}ta est:~* proptérea memor ero tui de terra Jordánis, et Hermóniim \textit{a} \textit{mon}\textit{te} \textbf{mó}\textbf{di}co.
	
	9. Abýssus a\textit{býs}\textit{sum} \textbf{ín}vocat,~* in voce catarac\textit{tá}\textit{rum} \textit{tu}\textbf{á}rum.
	
	10. Omnia excélsa tua, et \textit{fluc}\textit{tus} \textbf{tu}i~* super \textit{me} \textit{trans}\textit{i}\textbf{é}runt.
	
	11. In die mandávit Dóminus misericór\textit{di}\textit{am} \textbf{su}am:~* et nocte \textit{cán}\textit{ti}\textit{cum} \textbf{e}jus.
	
	12. Apud me orátio Deo \textit{vi}\textit{tæ} \textbf{me}æ,~* dicam Deo: \textit{Su}\textit{scép}\textit{tor} \textbf{me}\textbf{us} es.
	
	13. Quare oblí\textit{tus} \textit{es} \textbf{me}i?~* et quare contristátus incédo, dum afflígit \textit{me} \textit{in}\textit{i}\textbf{mí}cus?
	
	14. Dum confringúntur \textit{os}\textit{sa} \textbf{me}a,~* exprobravérunt mihi qui tríbulant me in\textit{i}\textit{mí}\textit{ci} \textbf{me}i.
	
	15. Dum dicunt mihi per síngulos dies: Ubi est \textit{De}\textit{us} \textbf{tu}us?~* quare tristis es, ánima mea? et \textit{qua}\textit{re} \textit{con}\textbf{túr}\textbf{bas} me?
	
	16. Spera in Deo, quóniam adhuc confi\textit{té}\textit{bor} \textbf{il}li:~* salutáre vultus mei, \textit{et} \textit{De}\textit{us} \textbf{me}us.
	
	17. Requiem æternam dona \textit{eis}, \textbf{Do}mine~* Et lux perpetua luc\textit{eat} \textbf{e}is.
 }{
 	1. As the hart panteth after the fountains of water; so my soul panteth after thee, O God.
 	
 	2. My soul hath thirsted after the strong living God; when shall I come and appear before the face of God?
 	
 	3. My tears have been my bread day and night, whilst it is said to me daily: Where is thy God?
 	
 	4.  These things I remembered, and poured out my soul in me: for I shall go over into the place of the wonderful tabernacle, even to the house of God:
 	
 	5. With the voice of joy and praise; the noise of one feasting.
 	
 	6. Why art thou sad, O my soul? and why dost thou trouble me?
 	
 	7. Hope in God, for I will still give praise to him: the salvation of my countenance,
 	
 	8. And my God. My soul is troubled within myself: therefore will I remember thee from the land of Jordan and Hermoniim, from the little hill.
 	
 	9. Deep calleth on deep, at the noise of thy flood-gates. 
 	
 	10. All thy heights and thy billows have passed over me.
 	
 	11. In the daytime the Lord hath commanded his mercy; and a canticle to him in the night. 
 	
 	12. With me is prayer to the God of my life. I will say to God: Thou art my support.
 	
 	13. Why hast thou forgotten me? and why go I mourning, whilst my enemy afflicteth me?
 	
 	14.  Whilst my bones are broken, my enemies who trouble me have reproached me.
 	
 	15. Whilst they say to me day be day: Where is thy God? Why art thou cast down, O my soul? and why dost thou disquiet me?
 	
 	16. Hope thou in God, for I will still give praise to him: the salvation of my countenance, and my God.
 	
 	17. Eternal rest, grant unto them O Lord, and let perpetual light shine upon them.
 } 
\end{latinenglishsection}
 
\gresetinitiallines{1}
\gregorioscore{antiphon_sitivit_anima}
\textit{My soul has thirsted after the living God: when shall I come and appear before the face of the Lord?}

\begin{latinenglishsection}
\latinenglish{
	{\color{red}\Vbar.} Ne tradas b\'{e}stis \'{a}nimas confit\'{e}ntes tibi.
	
	{\color{red}\Rbar.} Et \'{a}nimas p\'{a}uperum tu\'{o}rum ne oblivisc\'{a}ris in finem.
	
	Pater noster, {\color{red}\textit{secreto...}}
}{
	{\color{red}\Vbar.} Deliver not to beasts the souls that confess Thee.
	
	{\color{red}\Rbar.} And the souls of Thy poor, forget not to the end.
	
	Our Father, {\color{red}\textit{the rest in silence...}}
}
\end{latinenglishsection}

\vfill\pagebreak
 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Seventh Lesson - Job 17:1-3; 11-15}
 
\begin{latinenglishsection}
\latinenglish{
	Spiritus meus attenu\'{a}bitur, dies mei breviab\'{u}ntur, et solum mihi s\'{u}perest sep\'{u}lchrum. Non pecc\'{a}vi, et in amaritudi\'{i}nibus mor\'{a}tur \'{o}culus meus. L\'{i}bera me, D\'{o}min, et pone me juxta te, cuj\'{u}svis manus pugnet contra me. Dies mei transi\'{e}runt, cogitati\'{o}nes me\ae dissip\'{a}te sunt, torqu\'{e}ntes cor meum. Noctem vert\'{e}runt in diem, et rursum post t\'{e}nebras spero lucem. Si sustin\'{i}ero, inf\'{e}rnus domus mea est, et in t\'{e}nebris stravi l\'{e}ctulum meum. Putr\'{e}dini dixi: Pater meus es; mater mea et soror mea, v\'{e}rmibus. Ubi est ergo nunc pr\ae stol\'{a}tio mea, et pati\'{e}ntiam meam quis cons\'{i}derat?
	
	{\color{red}\Rbar.} Pecc\'{a}ntem me quot\'{i}die, et non me p\oe nit\'{e}ntem, timor mortis cont\'{u}rbat me: * Quia in inf\'{e}rno nulla est red\'{e}mptio, miser\'{e}re mei, Deus, et salva me.
	
	{\color{red}\Vbar.} Deus, in n\'{o}mine tuo salvum me fac, et in virt\'{u}te tua l\'{i}bera me. * Quia in inf\'{e}rno nulla est r\'{e}demptio, miser\'{e}re mei, Deus, et salva me.
}{
	My spirit is wasted: my days are shortened, and only the grave remaineth for me. I have not sinned, and my eye abideth in bitterness. Deliver me, O Lord, and set me beside Thee, and let any man's hand fight against me. My days have passed away; my thoughts are broken off, tormenting my heart. They have turned night into day; and after darkness I hope for light again. If I wait, hell is my house; and I have made my bed in darkness. I have said to rottenness: Thou art my father; to worms: You are my mother and my sister. Where then is now my expectation, and who considereth my patience?
	
	{\color{red}\Rbar.} Forasmuch as I sin daily, and repent not, the fear of death troubles me: * Because in hell there is no redemption, have mercy on me, O God, and save me.
	
	{\color{red}\Vbar.} O God, in Thy name save me, and in Thy power deliver me. * Because in hell there is no redemption, have mercy on me, O God, and save me.
}
\end{latinenglishsection}

\subsection*{Eighth Lesson - Job 19:20-27}
 
\begin{latinenglishsection}
\latinenglish{
	Pelli me\ae cons\'{u}mptis c\'{a}rnibus, adh\'{\ae}sit os meum, et derel\'{i}cta sunt tant\'{u}mmodo, l\'{a}bia circa dentes meos. Miser\'{e}mini mei, miser\'{e}mini mei, saltem vos, am\'{i}ci mei, qui manus D\'{o}mini t\'{e}tigit me. Quare persequi\'{i}mini me sicut Deus, et c\'{a}rnibus meis satur\'{a}min\'{i}? Quis mihi tr\'{i}buat ut scrib\'{a}ntur serm\'{o}nes mei? Quis mihi det ut exar\'{e}ntur in libro, stilo f\'{e}rreo et plumbi l\'{a}mina, vel celte sculp\'{a}ntur in s\'{i}lice? Scio enim quod red\'{e}mptor meus vivit, et in nov\'{i}ssimo die de terra surrect\'{u}rus sum: et rursum circ\'{u}mdabor pelle mea, et in carne mea vid\'{e}bo Deum meum. Quem vis\'{u}rus sum ego ipse, et \'{o}culi mei conspect\'{u}ri sunt, et non \'{a}lius: rep\'{o}sita est h\ae c spes mea in sinu meo.
	
	{\color{red}\Rbar.} D\'{o}mine, sec\'{u}ndum actum meum noli me judic\'{a}re: nihil dignum in consp\'{e}ctu tuo egi; ideo d\'{e}precor majest\'{a}tem tuam, * Ut tu, Deus, d\'{e}leas iniquit\'{a}tem meam.
	
	{\color{red}\Vbar.} Amplius lava me, D\'{o}mine, ab injust\'{i}tia mea, et a del\'{i}cto meo munda me. * Ut tu, Deus, d\'{e}leas inquit\'{a}tem meam.
}{
	The flesh being consumed, my bone cleaveth to my skin: and nothing but lips are left about my teeth. Have pity on me, have pity on me, at least ye my friends; because the hand of the Lord hath touched me. Why do ye persecute me as God, and glut yourselves with my flesh? Who will grant me that my words may by written? Who will grant me that they may be marked down in a book with an iron style and on a plate of lead, or else be graven with an instrument on the rock? For I know that my Redeemer liveth, and on the last day I shall rise out of the earth. And I shall be clothed again with my skin, and in my flesh I shall see my God. Whom I myself shall see, and my eyes shall behold, and not another: this, my hope, is laid up in my bosom.
	
	{\color{red}\Rbar.} Judge me not, O Lord, according to my deeds, for I have done nothing worthy in Thy sight; therefore I beseech Thy majesty, * that Thou, O God, mayst blot out my iniquity.
	
	{\color{red}\Vbar.} Wash me, O Lord, yet more from my injustice, and cleanse me from my sin. * That Thou, O God, mayst blot out my iniquity.
}
\end{latinenglishsection}

\subsection*{Ninth Lesson - Job 10:18-22}
 
\begin{latinenglishsection}
\latinenglish{
	Quare de vulva edux\'{i}sti me? qui \'{u}tinam cons\'{u}mptus essem ne \'{o}culus me vid\'{e}ret. Fu\'{i}ssem quasi non essem, de \'{u}tero transl\'{a}tus ad t\'{u}mulum. Numquid non p\'{a}ucitas di\'{e}rum me\'{o}rum fini\'{e}tur brevi? Dim\'{i}tte ergo me, ut plangam p\'{a}ululum dol\'{o}rem meum, \'{a}ntequam vadam, et non rev\'{e}rtar, terram mis\'{e}ri\ae et tenebr\'{a}rum, ubi umbra mortis et nullus ordo, sed sempit\'{e}rnus horror inh\'{a}bitat.
}{
	Why didst Thou bring me forth out of the womb? O that I had been consumed, that eye might not see me. I should have been as if I had not been carried from the womb to the grave. Shall not the fewness of my days be ended shortly? Suffer nme then that I may lament my sorrow a little; before I go and return no more, to a land that is dark and covered with the mist of death: a land of misery and darkness, where the shadow of death, and no order, but everlasting horror dwelleth.
}
\end{latinenglishsection}

{\color{red}\textit{The following responsory is said when only the third Nocturn has been said, Otherwise, it is omitted and the next rubric should be consulted.}}

\begin{latinenglishsection}
\latinenglish{
	{\color{red}\Rbar.} Libera me, D\'{o}mine, de viis inf\'{e}rni, qui portas \ae reas confreg\'{i}sti, et visit\'{a}sti inf\'{e}rnum, et ded\'{i}sti eis lumen, ut vid\'{e}rent te * Qui erant in p\oe nis tenebr\'{a}rum.
	
	{\color{red}\Vbar.} Clam\'{a}ntes dic\'{e}ntes: Adven\'{i}sti, Red\'{e}mptor noster. * Qui erant in p\oe nis tenebr\'{a}rum.
	
	{\color{red}\Vbar.} R\'{e}quiem \ae t\'{e}rnam dona eis, D\'{o}mine, et lux perp\'{e}tua l\'{u}ceat eis. * Qui erant in p\oe nis tenebr\'{a}rum.
}{
	{\color{red}\Rbar.} Deliver me, O Lord, from the ways of hell, Thou who hast broken the brazen gates, and hast visited hell, and hast given light to them, that they may behold Thee. * Who were in the pains of darkness.
	
	{\color{red}\Vbar.} Crying, and saying: Thou art come, O our Redeemer. * Who were in the pains of darkness.
	
	{\color{red}\Rbar.} Eternal rest grant unto them, O Lord, and let perpetual light shine upon them. * Who were in the pains of darkness.
}
\end{latinenglishsection}

{\color{red}\textit{The following Responsory takes the place of the preceding one when the three Nocturns have been said.}}

\begin{latinenglishsection}
\latinenglish{
	{\color{red}\Rbar.} L\'{i}bera me, D\'{o}mine, de morte \ae t\'{e}rna in die illa trem\'{e}nda, * Quando c\ae li mov\'{e}ndi sunt et terra: * Dum v\'{e}neris judic\'{a}re s\'{\ae}culum per ignem.
	
	{\color{red}\Vbar.} Tremens factus sum ego et t\'{i}meo, dum disc\'{u}ssio v\'{e}neris judic\'{a}re s\'{\ae}culum per ignem.
	
	{\color{red}\Vbar.} Dies illa, dies ir\ae, calamit\'{a}tis et mis\'{e}ri\ae, dies magna et am\'{a}ra valde. * Dum disc\'{u}ssio v\'{e}neris judic\'{a}re s\'{\ae}culum per ignem.
	
	{\color{red}\Vbar.} R\'{e}quiem \ae t\'{e}rnam dona eis, D\'{o}mine: et lux perp\'{e}tua l\'{u}ceat eis. L\'{i}bera me, D\'{o}mine, de morte \ae t\'{e}rna in die illa trem\'{e}nda. * Quando c\ae li mov\'{e}ndi sunt et terra. * Dum v\'{e}neris judic\'{a}re s\'{\ae}culum per ignem.
}{
	{\color{red}\Rbar.} Deliver me, O Lord, from the eternal death, in that dreadful day: * When the heavens and earth are to be moved: * When THou shalt come to judge the world by fire.
	
	{\color{red}\Vbar.} I tremble and do fear when the examination is to be, and Thy wrath to come. * When the heavens and earth are to be moved:
	
	{\color{red}\Vbar.} That day is the day of anger, of calamity, and of misery, a great day and very bitter. * When Thou shalt come to judge the world by fire.
	
	{\color{red}\Vbar.} Eternal rest grant unto them, O Lord, and let perpetual light shine upon them. Deliver me, O Lord, from eternal death, in that dreadful day, * When the heavens and earth are to be moved. * When Thou shalt come to judge the world by fire.
}
\end{latinenglishsection}

{\color{red}\textit{Here Lauds follows immediately. Or if Lauds are omitted, after the last Responsory the Pater Noster and the relevant prayer(s) on the following page is/are said.}}

\vfill\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Proper Collects for the Dead}

\subsection*{\color{red}Prayer on the Day of the Decease and Burial}

\begin{latinenglishsection}
\latinenglish{
	Absolve, qu\'{\ae}sumus, D\'{o}mine, \'{a}nimam f\'{a}muli tui (f\'{a}mul\ae\ tu\ae) {\color{red}N.}, ut def\'{u}nctus (def\'{u}ncta) s\'{\ae}culo tibi vivat: et, qu\ae per fragilit\'{a}tem carnis hum\'{a}na\\ conversati\'{o}ne comm\'{i}sit, tu v\'{e}nia misericord\'{i}ssim\ae\\ piet\'{a}tis abst\'{e}rge. Per D\'{o}minum nostrum Jesum Christum F\'{i}lium tuum, qui tecum vivit et regnat in unit\'{a}te Sp\'{i}ritus Sancti, Deus, per \'{o}mnia s\'{\ae}cula s\ae cul\'{o}rum.
	
	{\color{red}\Rbar.} Amen.
}{
	Absolve, we beseech Thee, O lord, the soul of Thy servant {\color{red}N.}, that being dead to the world he (or she) may live to Thee; and whatever he (or she) has committed through human frailty, do Thou wipe away by the pardon of Thy most merciful goodness. Through Our Lord Jesus Christ, Thy Son, who liveth and reigneth with Thee in the unity of the Holy Ghost, God, world without end.
	
	{\color{red}\Rbar.} Amen.
}
\end{latinenglishsection}

{\color{red}\textit{The following prayer may also be used:}}

\begin{latinenglishsection}
\latinenglish{
	Deus, cui pr\'{o}prium est miser\'{e}ri semper et p\'{a}rcere, te s\'{u}pplices exor\'{a}mus pro \'{a}nima f\'{a}muli tui {\color{red}N.} (f\'{a}mul\ae\ tu\ae {\color{red}N.}), quam h\'{o}die de hoc s\'{\ae}culo migr\'{a}re juss\'{i}sti: ut non tradas eam in manus inim\'{i}mici, neque oblivisc\'{a}ris in finem, sed j\'{u}beas eam a sanctis Angelis s\'{u}scipi, et ad p\'{a}triam parad\'{i}si perd\'{u}ci; ut, quia in te sper\'{a}vit et cr\'{e}diditm non p\oe nas inf\'{e}rni sust\'{i}neatm sed g\'{a}udia \ae t\'{e}rna poss\'{i}deat. Per Dóminum nostrum Iesum Christum Fílium tuum, qui tecum vivit et regnat in unitáte Spíritus Sancti, Deus, per ómnia s\'{\ae}cula sæculórum.
	
	{\color{red}\Rbar.} Amen.
}{
	O God, whose property it is always to have mercy and to spare, we humbly beseech Thee for the soul of Thy servant {\color{red}N.}, which today Thou hast taken out of this world: not to deliver it into the hands of the enemy, nor to forget it forever;but to command the holy angels to receive it, and to conduct it to the heavenly country; that since it has believed and hoped in Thee, it may not suffer the pains of hell, but possess eternal joys. Through Our Lord Jesus Christ Thy Son, who liveth and reigneth with Thee in the unity of the Holy Ghost, God, world without end.
	
	{\color{red}\Rbar.} Amen.
}
\end{latinenglishsection}

\subsection*{\color{red}Prayer for the Third, Seventh, or Thirtieth Day after Burial}

\begin{latinenglishsection}
\latinenglish{
	Qu\ae sumus, D\'{o}mine, ut \'{a}nimam f\'{a}muli tui {\color{red}N.},\\ (f\'{a}mul\ae\ tu\ae\ {\color{red}N.}), cujus depositi\'{o}nis diem t\'{e}rtium ({\color{red}vel} s\'{e}ptimum, {\color{red}vel} trig\'{e}simum) commemor\'{a}mus, sanct\'{o}rum atque elect'{o}rum tu\'{o}rum larg\'{i}ri dign\'{e}ris cons\'{o}rtium: et rorem miseric\'{o}rdi\ae tu\ae per\'{e}nnem inf\'{u}ndas. Per Dóminum nostrum Iesum Christum Fílium tuum, qui tecum vivit et regnat in unitáte Spíritus Sancti, Deus, per ómnia s\'{\ae}cula sæculórum.
	
	{\color{red}\Rbar.} Amen.
}{
	Vouchsafe, we beseech Thee, O Lord, to admit the soul of Thy servant {\color{red}N.}, the third (seventh or thirtieth) day of whose burial we commemorate, into the fellowship of Thy saints and elect, and sprinkle upon it the perpetual dew of Thy mercy. Through Our Lord Jesus Christ Thy Son, who liveth and reigneth with Thee in the unity of the Holy Ghost, God, world without end.
	
	{\color{red}\Rbar.} Amen.
}
\end{latinenglishsection}

\subsection*{\color{red}Prayer on the Anniversary of Burial}

\begin{latinenglishsection}
\latinenglish{
	Deus, indulgenti\'{a}rum D\'{o}mine, da \'{a}nim\ae\ f\'{a}muli tui {\color{red}N.},\\ (f\'{a}mul\ae\ tu\ae\ {\color{red}N.}, {\color{red}vel} anim\'{a}bus famul\'{o}rum famalur\'{u}mque tu\'{a}rum), cujus (quorum) annivers\'{a}rium depositi\'{o}nis diem commemor\'{a}mus, refrig\'{e}rii sedem, qui\'{e}tis\\ beatit\'{u}dinem, et l\'{u}minis clarit\'{a}tem. Per Dóminum nostrum Iesum Christum Fílium tuum, qui tecum vivit et regnat in unitáte Spíritus Sancti, Deus, per ómnia s\'{\ae}cula sæculórum.
	
	{\color{red}\Rbar.} Amen.
}{
	O God, the Lord of mercy and pardon, grant to the soul of Thy servant {\color{red}N.} ({\color{red}or} servants), the anniversary of whose burial we commemorate, the place of refreshment, the blessedness of rest, and the brightness of light. Through Our Lord Jesus Christ Thy Son, who liveth and reigneth with Thee in the unity of the Holy Ghost, God, world without end.
	
	{\color{red}\Rbar.} Amen.
}
\end{latinenglishsection}

\vfill\pagebreak

\subsection*{\color{red}For a Deceased Pope}

\begin{latinenglishsection}
\latinenglish{
	Deus, qui inter summos sacerd\'{o}tes f\'{a}mulum tuum {\color{red}N.} ineff\'{a}bili tua dispositi\'{o}ne connumer\'{a}ri volu\'{i}sti: pr\ae sta, qu\'{\ae}sumus; ut qui Unig\'{e}niti F\'{i}lii tui vices in terris ger\'{e}bat, sanct\'{o}rum tu\'{o}rum Pont\'{i}ficum cons\'{o}rtio perp'{e}tuo aggreg\'{e}tur. Per e\'{u}ndem D\'{o}minum nostrum Iesum Christum Fílium tuum, qui tecum vivit et regnat in unitáte Spíritus Sancti, Deus, per ómnia s\'{\ae}cula sæculórum.
	
	{\color{red}\Rbar.} Amen.
}{
	O God, by whose favor Thy servant {\color{red}N.} was raised to the dignity of the sovereign pontiff, grant, we beseech Thee, that he may be admitted to the eternal society of Thy holy pontiffs in heaven. Through the same Our Lord Jesus Christ Thy Son, who liveth and reigneth with Thee in the unity of the Holy Ghost, God, world without end.
	
	{\color{red}\Rbar.} Amen.
}
\end{latinenglishsection}

\begin{center}
\par\noindent\rule{0.5\textwidth}{0.4pt}
\end{center}

\subsection*{\color{red}For a Deceased Bishop}

\begin{latinenglishsection}
\latinenglish{
	Deus, qui inter apost\'{o}licos sacerd\'{o}tes f\'{a}mulum tuum {\color{red}N.} (f\'{a}mulos tuos {\color{red}N.} et {\color{red}N.}) pontific\'{a}lis fec\'{i}sti dignit\'{a}te vig\'{e}re: pr\ae sta, qu\'{\ae}sumus, ut e\'{o}rum quoque perp\'{e}tuo aggreg\'{e}(n)tur cons\'{o}rtio. Per D\'{o}minum nostrum Iesum Christum Fílium tuum, qui tecum vivit et regnat in unitáte Spíritus Sancti, Deus, per ómnia s\'{\ae}cula sæculórum.
	
	{\color{red}\Rbar.} Amen.
}{
	O God, by whose favor Thy servant(s) {\color{red}N.} (and {\color{red}N.}) was raised to the dignity of a bishop, and honored with the apostolic functions, grant, we beseech Thee, that he (they) may be admitted to the eternal society of Thy apostles in heaven. Through the same Lord Jesus Christ Thy Son, who liveth and reigneth with Thee in the unity of the Holy Ghost, God, world without end.
	
	{\color{red}\Rbar.} Amen.
}
\end{latinenglishsection}

\subsubsection*{\color{red}For a Cardinal-Bishop, say:}

\begin{latinenglishsection}
\latinenglish{
	f\'{a}mulum tuum {\color{red}N.} Ep\'{i}scopum Cardin\'{a}lem pontific\'{a}li fec\'{i}sti dign\'{i}tate vig\'{e}re.
}{
	O God, by whose favor Thy servant {\color{red}N.} was raised to the dignity of Cardinal Bishop.
}
\end{latinenglishsection}

\subsubsection*{\color{red}For a Cardinal-Priest or Cardinal-Deacon who was also a Bishop, say:}

\begin{latinenglishsection}
\latinenglish{
	f\'{a}mulum tuum {\color{red}N.} Presb\'{y}terum ({\color{red}vel} Diaconum)\\ Cardin\'{a}lem pontific\'{a}li fec\'{i}sti dignit\'{a}te vig\'{e}re.
}{
	O God, by whose favor Thy servant {\color{red}N.}. was raised to the dignity of Cardinal Priest.
}
\end{latinenglishsection}

\subsubsection*{\color{red}For a Cardinal-Priest who was not a Bishop, say:}
\begin{latinenglishsection}
\latinenglish{
	f\'{a}mulum tuum {\color{red}N.} Presb\'{y}terum Cardin\'{a}lem sacerdot\'{a}li fec\'{i}sti dignit\'{a}te vig\'{e}re.
}{
	O God, by whose favor Thy servant {\color{red}N.}. was raised to the dignity of Cardinal Priest.
}
\end{latinenglishsection}

\subsubsection*{\color{red}For a Cardinal-Deacon who was a priest, say:}
\begin{latinenglishsection}
\latinenglish{
	f\'{a}mulum tuum {\color{red}N.} Di\'{a}conum Cardin\'{a}lem sacerdot\'{a}li fec\'{i}sti dignit\'{a}te vig\'{e}re.
}{
	O God, by whose favor Thy servant {\color{red}N.}. was raised to the dignity of Cardinal Deacon.
}
\end{latinenglishsection}

\begin{center}
\par\noindent\rule{0.5\textwidth}{0.4pt}
\end{center}



\end{document}